Joint estimation of phylogenies and ancestral states is not consistent

### 2017-12-26 appendix.tex
noticed minor error; fix plus add clarification

\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$.

@@

See Appendix for full calculations.}

@@

Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is

@@

    &\qquad - \log(8) \nonumber \\

@@

We use the facts that, for $x,y\in[0,1]$,

@@

1-x^2-y^2-x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y) \\

@@

4xy & = 2x \cdot 2y \le (1+x^2)(1+y^2) \le (1+x)(1+y).

@@

The $p_{\emptyset}$ term is bounded as

@@

+ p_{123}        \cdot\log((1+x)(1+y)(1-y))
- \log(8) \\

@@

+ p_{13}\cdot\log(1-w)
- \log(8) \\

@@

+ p_{13}\cdot\log(1-w)
- \log(8).

@@

Trivially $0 \le \hat{w} \le 1$.
We check that $0 \le \hat{x} \le 1$ and $0 \le \hat{y} \le 1$ so these maximizers are valid.

@@

we have
\hat{x} = \frac{3-\tilde{p} - 2\tilde{p}}{3-\tilde{p}+2\tilde{p}} = \frac{3-3\tilde{p}}{3+\tilde{p}},
and, since $0 \le \tilde{p} \le 1/2$, $0 \le \hat{x} \le 1$.

@@

+ p_{13}\cdot\log(2p_{13})
- \log(8) \\

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
%EM Yay! Would be even better if you specified which upper bound (eqref) you want us to be looking at... there are many.
For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
%das: not quite sure what this comment means actually.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
%das: is this a better explanation or did you want it written out explicitly?
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2)
- \log(8) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2)
- \log(8) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

We now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
It will suffice to show
$$
0 \le \tilde{p} - 2p_{13} \le 1
$$
and
$$
0 \le \tilde{p} + 2p_{13} \le 1
$$
for $\tilde{p}$ defined above, which is equivalent to showing $p_{13} \le 1/4$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right).
$$
We see that
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 2 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - (x^*)^2.
\label{eq:lenny}
\end{equation}
Note
$$
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 1 - 4x^* + (x^*)^2;
$$
showing that \eqref{eq:lenny} holds if $x^* \le 2$, which is satisfied by definition.
Thus, $p_{13} \le 2/8 = 1/4$ and our $\hat{x}$ and $\hat{y}$ are valid.

@@

The full likelihood in this case is then bounded below by

@@

+ b      \cdot\log(b)

@@

i.e.\ the maximum is achieved at $\{1,1-b,0\}$.

@@

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
$$
\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
$$
where
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative of the log-likelihood with respect to $w$ is
$$
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
Using the quadratic formula with
$$
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
$$
which is a function of the generating parameters $x^*$ and $y^*$.
We see that
$$
\hat{w} \ge 1
$$
is equivalent to
%-b - \sqrt{b^2 - 4ac} \le 2a,
%-b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
$$
a + b + c \ge 0,
$$
which, using
$$
\alpha_1 + \alpha_2 = \frac{1}{2}\beta
$$
and
$$
\alpha_1 - \alpha_2 = x^*(y^*)^2,
$$
and simplifying as functions of $\gamma$ and $\beta$ yields
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
\end{equation}
\end{proof}

>>>>

\caption{Likelihood calculations for all site patterns $\siteSplit_j$ and ancestral state partitions $\ancestralSplitPartition_j$ after maximizing over ancestral states.

@@

See Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} for full calculations.}

@@

Directly substituting calculations from Table~\ref{tab:sitepatprob} and Table~\ref{tab:likelihoods} where $\tau = \tau_1$ in \eqref{eq:site_pattern_profile_likelihood_mean}, the log-likelihood is, suppressing normalizing constants $-\log 8$ and $-\log 32$,

@@

@@

Our plan is to simplify the likelihood to remove log-of-quadratic terms of $x,y,$ and $w$, obtaining a likelihood that has a closed-form maximum in each variable.
To bound the generating probabilities, we use the facts that, for $x,y\in[0,1]$,

@@

p_{12} = p_{23} = 1-x^2-y^2+x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
p_{1} = p_{3} = 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
p_{2} = p_{123} = 1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y).
To bound the remaining $p_{\emptyset}$ and $p_{13}$, we use

@@

4xy & = 2x \cdot 2y \le (1+x^2)(1+y^2) \le (1+x)(1+y)

@@

so that the $p_{\emptyset}$ term is bounded as

@@

+ p_{123}        \cdot\log((1+x)(1+y)(1-y)) \\

@@

+ p_{13}\cdot\log(1-w) \\

@@

+ p_{13}\cdot\log(1-w).

@@

Trivially $0 \le \hat{w} \le 1$ and $\hat{x}, \hat{y} \le 1$.
We check that $\hat{x}, \hat{y} \ge 0$ so these maximizers are valid.
For $\hat{x} \ge 0$ we need $a_1 \ge a_2$; letting

@@

we have from \eqref{eq:a_const} that $a_1 \ge a_2$ only if $3 \ge 3\tilde{p}$, and, since $0 \le \tilde{p} \le 1/2$, $\hat{x} \ge 0$.

@@

+ p_{13}\cdot\log(2p_{13}) \\

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
The upper bound in \eqref{eq:farris-upper-bound} is for the ancestral state split $\{1\}$.
%EM Yay! Would be even better if you specified which upper bound (eqref) you want us to be looking at... there are many.
For the ancestral state split $\emptyset$ we see the upper bound will be the same except we lose a $(1+x)^2$ term, gain a $(1-x)^2$ term, lose a $(1-w)$ term, and gain a $(1+w)$ term.
In this case, all terms involving $w$ simplify to $\log(1+w)$, maximized at $\hat{w}=1$.
For the constants above, $a_{1}$ is the multiplier for the $\log(1+x)$ term and $a_{2}$ for the $\log(1-x)$ term meaning that exchanging the above quadratic $x$ terms yields new constants
%EM It would seem helful for the below check to express these a_1' etc in terms of tilde{p}.
%das: not quite sure what this comment means actually.
\begin{equation}
    \begin{aligned}
        a_{1}' &= a_{1}-2p_{13}, \\
        a_{2}' &= a_{2}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_x}
\end{equation}
for site split $\emptyset$ and
\begin{equation}
    \begin{aligned}
        a_{3}' &= a_{3}-2p_{13}, \\
        a_{4}' &= a_{4}+2p_{13}, \\
    \end{aligned}
    \label{eq:a_const_prime_y}
\end{equation}
for site split $\{1,2\}$.
%EM This check would make a lot more sense if you put the C^{2}_{\tau_1,\tau_1}(x^*, y^*)  and the other analogous definition first.
%EM It would also be nice to do one of the calculations to get us from a_1 world to a_1' world. Just staring at table S4 doesn't make this obvious.
%das: is this a better explanation or did you want it written out explicitly?
To bound the case of $\ancestralSplitPartition_j(\tau_1, t) = \emptyset$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'}
+ a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
&\qquad\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}}
+ a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}}
+ \log(2) \\
&\qquad := C^{2}_{\tau_1,\tau_1}(x^*, y^*),
\end{align*}
and, for the case of $\ancestralSplitPartition_j(\tau_1, t) = \{1,2\}$,
\begin{align*}
&    \ell_{\tau_1,t^*}(\tau_1, t) \\
&\qquad\le      p_{\emptyset}  \cdot\log(2)
+ a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}}
+ a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
&\qquad\qquad+ a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'}
+ a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'}
+ \log(2) \\
&\qquad := C^{3}_{\tau_1,\tau_1}(x^*, y^*).
\end{align*}

We now need to check that
$$
\hat{x} = \frac{a_{1}'-a_{2}'}{a_{1}'+a_{2}'}, \ \hat{y} = \frac{a_{3}'-a_{4}'}{a_{3}'+a_{4}'}
$$
are both between zero and one.
Using a similar argument as that where $\ancestralSplitPartition_j(\tau_1, t) = \{1,3\}$, we only need to show $a_1' \ge a_2'$, which is true if $3-2p_{13} \ge 3\tilde{p}+2p_{13}$.
Some rearranging and the fact that $0 \le \tilde{p} \le 1/2$ shows this is equivalent to showing $p_{13} \le 3/8$.
Using Table~\ref{tab:sitepatprob} and $t^*$,
$$
p_{13} = \frac{1}{8}\left(1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2\right),
$$
meaning we are interested in whether
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3.
\label{eq:generating_ineq}
\end{equation}
We see that
\begin{equation}
1 + (x^*)^2 + (y^*)^2 - 4x^*(y^*)^2 + (x^*)^2(y^*)^2 \le 3 \iff (y^*)^2\left(1 - 4x^* + (x^*)^2\right) \le 2 - (x^*)^2.
\label{eq:lenny}
\end{equation}
and that
\begin{align*}
(y^*)^2\left(1 - 4x^* + (x^*)^2\right) &\le (y^*)^2\left(1 - 2x^* + (x^*)^2\right) \\
                                       &=   (y^*)^2\left(1 - x^*\right)^2 \\
                                       &\le        \left(1 - x^*\right)^2 \\
                                       &\le              2 - (x^*)^2,
\end{align*}
with the last inequality holding if $0 \le x^* \le 1$, showing that \eqref{eq:lenny} holds.
Thus, $p_{13} \le 3/8$ and our $\hat{x}$ and $\hat{y}$ are valid.

@@

The partial likelihood in this case is then bounded below by

@@

+ b      \cdot\log(b),

@@

i.e., the maximum is achieved at $\{1,1-b,0\}$.

@@

Now substitute in values for $p_{\emptyset}$ and $p_{13}$ from Table~\ref{tab:sitepatprob}.
Simplifying, the likelihood can be written
$$
\ell(w) = \alpha_1\log(\beta+\gamma w) + \alpha_2\log(\beta-\gamma w) + \log(1+w)
$$
where
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
Since $x^*$ and $y^*$ fall between zero and one, we make use of the inequalities (excepting the cases of $x^*=y^*=0$ and $x^*=y^*=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative of the log-likelihood with respect to $w$ is
$$
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{\alpha_1 \gamma}{\beta+\gamma w} - \frac{\alpha_2 \gamma}{\beta-\gamma w} + \frac{1}{1+w}.
$$
The inequality $\beta > \gamma$ implies that this function stays finite and that, when considering $\ell'(w) \lessgtr 0$, we equivalently consider $f(w) \lessgtr 0$ where $f$ is the quadratic function
\begin{align*}
f(w) &= (w)^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2).
\end{align*}
% Note from das: b^2 - 4ac is always \ge 0 by \alpha_1 > \alpha_2, so b^2 - 4ac can only be zero if b is zero and a or c is zero. c is never zero, so this means a must be zero for this to happen, which only occurs if \gamma = 0, i.e., if x or y is zero.
Excepting all the cases where $x^*=0$ or where $y^*=0$, this implies $\ell'$ has two zeros according to the quadratic formula.
Because $\alpha_1 > \alpha_2$ we have $\ell'(0) > 0$, and thus $\ell$ is increasing at $w=0$.
This implies that if the smaller of the zeros of $f(w)$ is greater than one, then $\hat{w} \equiv 1$.
Using the quadratic formula with
$$
a = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w} = \frac{-b - \sqrt{b^2 - 4ac}}{2a},
$$
which is a function of the generating parameters $x^*$ and $y^*$.
We see that $a \le 0$ and
$$
\hat{w} \ge 1
$$
is equivalent to
%-b - \sqrt{b^2 - 4ac} \le 2a,
%0 \ge -b - 2a \le \sqrt{b^2 - 4ac},
%b^2 + 4ab + 4a^2 \le b^2 - 4ac,
%a^2 + ab + ac \le 0,
$$
a + b + c \ge 0,
$$
which, using
$$
\alpha_1 + \alpha_2 = \frac{1}{2}\beta
$$
and
$$
\alpha_1 - \alpha_2 = x^*(y^*)^2,
$$
and simplifying as functions of $\gamma$ and $\beta$ yields
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0.
\end{equation}
\end{proof}
